Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-24 numerical-gravity.tex
Jeff Oishi's changes.
Changed all "X order" to "x-order" and "finite-difference" and "finite-volume" to "finite difference" and "finite volume", to ensure consistency.

To compute the potential field from this gravitating mass field on the root grid, we use a a fast Fourier transform. For periodic boundary conditions, we can use either a simple Greens function kernel of $-k^{-2}$, or the finite-difference equivalent \citep{Hockney88}:
\begin{equation}
G(\vec{k}) = - \frac{\Delta x}{2 \left( \sin(k_x \Delta x/2)^2 + \sin(k_y \Delta y/2)^2 + \sin(k_z \Delta z/2)^2 \right) }
\end{equation}
where $k^2 = k_x^2 + k_y^2 + k_z^2$ is the wavenumber in Fourier space and the potential is calculated in k-space as usual with $\tilde{\phi}(k) = G(k) \tilde{\rho}(k)$.

@@

In order to calculate more accurate potentials on the subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub-grids as before.  To compute the potential on subgrids, however, we use the standard seven-point (in three dimensions) second-order finite-difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results but the default is the tri-linear interpolation, which empirically provides a resonable compromise between speed and accuracy. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each subgrid separately.

>>>>

To compute the potential field from this gravitating mass field on the root grid, we use a a fast Fourier transform. For periodic boundary conditions, we can use either a simple Greens function kernel of $-k^{-2}$, or the finite difference equivalent \citep{Hockney88}:
\begin{equation}
G(\myvec{k}) = - \frac{\Delta x}{2 \left( \sin(k_x \Delta x/2)^2 + \sin(k_y \Delta y/2)^2 + \sin(k_z \Delta z/2)^2 \right) }
\end{equation}
where $k^2 = k_x^2 + k_y^2 + k_z^2$ is the wavenumber in Fourier space and the potential is calculated in k-space as usual with $\tilde{\phi}(k) = G(k) \tilde{\rho}(k)$.

@@

In order to calculate more accurate potentials on the subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub-grids as before.  To compute the potential on subgrids, however, we use the standard seven-point (in three dimensions) second-order finite difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results but the default is the tri-linear interpolation, which empirically provides a resonable compromise between speed and accuracy. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each subgrid separately.

### 2013-06-24 numerical-mhd-ct.tex
Changed all "X order" to "x-order" and "finite-difference" and "finite-volume" to "finite difference" and "finite volume", to ensure consistency.

The base Godunov method, described in \citet{Li08a}, uses spatially and
temporally second order reconstruction (both MUSCL-Hancock and
Piecewise Linear Method), and a selection of Riemann solvers including
HLLC and HLLD \citep{Mignone07}, as described earlier.  
The constrained transport methods
are the first-order method described by \citet{Balsara99} and the
second order methods described in \citet{Gardiner05}.  The AMR
machinery is described by \citet{Balsara01} and \citet{Collins10}.

>>>>

The base Godunov method, described in \citet{Li08a}, uses spatially and
temporally second-order reconstruction (both MUSCL-Hancock and
Piecewise Linear Method), and a selection of Riemann solvers including
HLLC and HLLD \citep{Mignone07}, as described earlier.  
The constrained transport methods
are the first-order method described by \citet{Balsara99} and the
second-order methods described in \citet{Gardiner05}.  The AMR
machinery is described by \citet{Balsara01} and \citet{Collins10}.

### 2013-06-24 numerical-overview.tex
Changed all "X order" to "x-order" and "finite-difference" and "finite-volume" to "finite difference" and "finite volume", to ensure consistency.

These equations are finite-differenced and for simplicity are solved with the
same timestep as the hydrodynamic equations, as discussed in 
Section~\ref{sec.ov.amr}.
The dark matter particles are distributed onto the grids
using the cloud-in-cell (CIC) interpolation technique to form a spatially
discretized density field (analogous to the baryon densities used to calculate
the equations of hydrodynamics).  After sampling dark matter density onto the
grid and adding baryon density if it exists (to get the total matter density
in each cell), the gravitational potential is calculated on the periodic root
grid using a fast Fourier transform.  In order to calculate more accurate
potentials on subgrids, \enzo\ re-samples the dark matter density onto
individual subgrids using the same CIC method as on the root grid, but at
higher spatial resolution (and again adding baryon densities if applicable).
Boundary conditions are then interpolated from the potential values on the
parent grid (with adjacent grid patches on a given level communicating to
ensure that their boundary values are the same), and then a multigrid
relaxation technique is used to calculate the gravitational potential at every
point within a subgrid.  Forces are computed on the mesh by
finite-differencing potential values and are then interpolated to the particle
positions, where they are used to update the particle's position and velocity
information.  Potentials on child grids are computed recursively and particle
positions are updated using the same timestep as in the hydrodynamic
equations.  Particles are stored in the most highly refined grid patch at the
point in space where they exist, and particles which move out of a subgrid
patch are sent to the grid patch covering the adjacent volume with the finest
spatial resolution, which may be of the same spatial resolution, coarser, or finer
than the grid patch that the particles are moved from.  This takes place in a
communication process at the end of each timestep on a level.

@@

PPM is an explicit, higher order-accurate version of
Godunov's method for ideal gas dynamics with third order-accurate piecewise parabolic
monotonic interpolation and a nonlinear Riemann solver for shock
capturing.  It advances the hydrodynamic equations in the following
steps:
\begin{enumerate}
 \item Construct monotonic parabolic interpolation of cell average
 data, for each fluid quantity.
 \item Compute interface states by averaging the parabola for over the domain of dependence for
 each interface
 \item Use interface data to solve the Riemann problem.
 \item Difference the interface fluxes to update the cell average quantities.
\end{enumerate}
It does an excellent job of capturing strong shocks in at
most two cells.  Multidimensional schemes are built up by directional
splitting and produce a method that is formally second order-accurate
in space and time which explicitly conserves mass, linear momentum, and energy.

@@

As a check on PPM, \enzo\ also includes an implementation of the
finite-difference hydrodynamic algorithm employed in the compressible 
magnetohydrodynamics code `\zeus'
\citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian
grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} 
within a multistep 
(operator split) solution procedure which is fully explicit in time.  
This method is formally second order-accurate in space but 
first order-accurate in time.  It is important to note that 
this method conserves internal energy
rather than total energy, so the ``dual energy
formulation'' discussed in Section~\ref{sec.ov.hydro.ppm} is unnecessary.

@@

The \zeus\ method uses a von Neumann-Richtmeyer artificial viscosity 
to smooth shock discontinuities that may
appear in fluid flows and can cause a break-down of finite-difference
equations.  The artificial viscosity term is added in the source terms
as
\begin{eqnarray}
\rho \frac{\partial\textbf{v}}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \textbf{v} - \textbf{Q} : \grad \textbf{v}, 
\end{eqnarray}
where \textbf{v} is the baryon velocity, $\rho$ is the mass density,
$p$ is the pressure, $e$ is the internal energy density of the gas and \textbf{Q}
is the artificial viscosity stress tensor, such that:

>>>>

These equations are finite differenced and for simplicity are solved with the
same timestep as the hydrodynamic equations, as discussed in 
Section~\ref{sec.ov.amr}.
The dark matter particles are distributed onto the grids
using the cloud-in-cell (CIC) interpolation technique to form a spatially
discretized density field (analogous to the baryon densities used to calculate
the equations of hydrodynamics).  After sampling dark matter density onto the
grid and adding baryon density if it exists (to get the total matter density
in each cell), the gravitational potential is calculated on the periodic root
grid using a fast Fourier transform.  In order to calculate more accurate
potentials on subgrids, \enzo\ re-samples the dark matter density onto
individual subgrids using the same CIC method as on the root grid, but at
higher spatial resolution (and again adding baryon densities if applicable).
Boundary conditions are then interpolated from the potential values on the
parent grid (with adjacent grid patches on a given level communicating to
ensure that their boundary values are the same), and then a multigrid
relaxation technique is used to calculate the gravitational potential at every
point within a subgrid.  Forces are computed on the mesh by
finite differencing potential values and are then interpolated to the particle
positions, where they are used to update the particle's position and velocity
information.  Potentials on child grids are computed recursively and particle
positions are updated using the same timestep as in the hydrodynamic
equations.  Particles are stored in the most highly refined grid patch at the
point in space where they exist, and particles which move out of a subgrid
patch are sent to the grid patch covering the adjacent volume with the finest
spatial resolution, which may be of the same spatial resolution, coarser, or finer
than the grid patch that the particles are moved from.  This takes place in a
communication process at the end of each timestep on a level.

@@

PPM is an explicit, higher order-accurate version of
Godunov's method for ideal gas dynamics with third-order-accurate piecewise parabolic
monotonic interpolation and a nonlinear Riemann solver for shock
capturing.  It advances the hydrodynamic equations in the following
steps:
\begin{enumerate}
 \item Construct monotonic parabolic interpolation of cell average
 data, for each fluid quantity.
 \item Compute interface states by averaging the parabola for over the domain of dependence for
 each interface
 \item Use interface data to solve the Riemann problem.
 \item Difference the interface fluxes to update the cell average quantities.
\end{enumerate}
It does an excellent job of capturing strong shocks in at
most two cells.  Multidimensional schemes are built up by directional
splitting and produce a method that is formally second-order-accurate
in space and time which explicitly conserves mass, linear momentum, and energy.

@@

As a check on PPM, \enzo\ also includes an implementation of the
finite difference hydrodynamic algorithm employed in the compressible 
magnetohydrodynamics code `\zeus'
\citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian
grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} 
within a multistep 
(operator split) solution procedure which is fully explicit in time.  
This method is formally second-order-accurate in space but 
first-order-accurate in time.  It is important to note that 
this method conserves internal energy
rather than total energy, so the ``dual energy
formulation'' discussed in Section~\ref{sec.ov.hydro.ppm} is unnecessary.

@@

The \zeus\ method uses a von Neumann-Richtmeyer artificial viscosity 
to smooth shock discontinuities that may
appear in fluid flows and can cause a breakdown of finite difference
equations.  The artificial viscosity term is added in the source terms
as
\begin{eqnarray}
\rho \frac{\partial\textbf{v}}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \textbf{v} - \textbf{Q} : \grad \textbf{v}, 
\end{eqnarray}
where \textbf{v} is the baryon velocity, $\rho$ is the mass density,
$p$ is the pressure, $e$ is the internal energy density of the gas and \textbf{Q}
is the artificial viscosity stress tensor, such that:

### 2013-06-24 numerical-ppm.tex
Changed all "X order" to "x-order" and "finite-difference" and "finite-volume" to "finite difference" and "finite volume", to ensure consistency.

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

With these effective states, an approximation to the Riemann problem is found
(see below for more detail about the Riemann solvers used), producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third order
accurate in space and second order accurate in time.  These are then
used to solve the difference Equations (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

@@

We note that the acceleration required in Equation~(\ref{eq:momentum_diff}) is actually the
acceleration felt by the entire zone and not just at the zone center.
Therefore, it is possible to find the mass weighted average acceleration over the zone
by expanding the density and acceleration distributions and
retaining all terms up to second order in $\Delta x$:
%
\begin{equation}
g_j^{n+1/2} = 
       \frac{1}{2 a^{n+1/2} \delta x_j} \left[ 
             \phi_{j+1}^{n+1/2} 
           - \phi_{j-1}^{n+1/2} 
           + \frac{1}{12} \left(    \phi^{n+1/2}_{j+1} 
                                 - 2\phi^{n+1/2}_j 
                                 + \phi^{n+1/2}_{j-1} \right) 
                                   {\delta d_j \over d_j}
       \right].
       \label{eq:accel}
\end{equation}
Generally we find that the potential is so slowly varying that this change is unnecessary.

@@

On rare occasions, high order solutions can cause negative densities or energies.  
Both our PPM and MUSCL solvers employ a Riemann
solver fallback mechanism \citep{Lemaster09}.  If a negative density is found at
a particular interface, the more diffusive HLL Riemann solver is used to compute the fluxes associated with that cell, and the flux update is repeated.

>>>>

We first construct monotonic piecewise 
parabolic (third-order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first-order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

With these effective states, an approximation to the Riemann problem is found
(see below for more detail about the Riemann solvers used), producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third-order
accurate in space and second-order accurate in time.  These are then
used to solve the difference Equations (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

@@

We note that the acceleration required in Equation~(\ref{eq:momentum_diff}) is actually the
acceleration felt by the entire zone and not just at the zone center.
Therefore, it is possible to find the mass weighted average acceleration over the zone
by expanding the density and acceleration distributions and
retaining all terms up to second-order in $\Delta x$:
%
\begin{equation}
g_j^{n+1/2} = 
       \frac{1}{2 a^{n+1/2} \delta x_j} \left[ 
             \phi_{j+1}^{n+1/2} 
           - \phi_{j-1}^{n+1/2} 
           + \frac{1}{12} \left(    \phi^{n+1/2}_{j+1} 
                                 - 2\phi^{n+1/2}_j 
                                 + \phi^{n+1/2}_{j-1} \right) 
                                   {\delta d_j \over d_j}
       \right].
       \label{eq:accel}
\end{equation}
Generally we find that the potential is so slowly varying that this change is unnecessary.

@@

On rare occasions, high-order solutions can cause negative densities or energies.  
Both our PPM and MUSCL solvers employ a Riemann
solver fallback mechanism \citep{Lemaster09}.  If a negative density is found at
a particular interface, the more diffusive HLL Riemann solver is used to compute the fluxes associated with that cell, and the flux update is repeated.

### 2013-06-24 numerical-radiation-fld.tex
Jeff Oishi's changes.
Mike Norman's edits.
Changed all "X order" to "x-order" and "finite-difference" and "finite-volume" to "finite difference" and "finite volume", to ensure consistency.

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR, non-SMR) grids, which has been tuned for large-scale simulations
involving many ionizing sources.  Detailed explanations of the model and
solution approach may be found in
\cite{NBHBROW2007}, \cite{ReynoldsHayesPaschosNorman2009} and
\cite{NRS2009}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.  
\enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities.
%\begin{eqnarray}
%  \label{eq:fld_radiation}
%  \partial_t E_r + \frac1a \div \left(E_r \vec{v}_b\right) &=& 
%    \div\left(D\grad E_r\right) -
%    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
%  \label{eq:fld_heating}
%  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \vec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
%\end{eqnarray}
%where $E_r$ is a grey radiation energy density, $e_c$ is the
%internal energy correction due to photo-heating and chemical cooling,
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  Here, we define the grey radiation energy density
%through first assuming a fixed frequency spectrum, i.e.
%$E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t) \chi(\nu)$, and then
%defining the integrated quantity
%\begin{equation}
%\label{eq:grey_radiation_energy}
%   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
%   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
%   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
%\end{equation}
%In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
%square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
%opacity, $\eta$ is the field of radiation sources, $G$ provides the
%radiation induced photo-heating, $\Lambda$ provides the chemical
%cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_heating}) along with the chemical network 
(Equation~\ref{eq:species_evolution}) is solved
independently of \enzo's hydrodynamics, gravity and dark-matter solvers
(Sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of Equations (\ref{eq:fld_radiation}) and
(\ref{eq:species_evolution}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given timestep to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite-difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

>>>>

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR, non-SMR) grids, which has been tuned for large-scale simulations
involving many ionizing sources.  Detailed explanations of the model and
solution approach may be found in
\citet{NBHBROW2007}, \citet{ReynoldsHayesPaschosNorman2009},
\citet{NRS2009}, and \citet{2013arXiv1306.0645N}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.  
\enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities.
%\begin{eqnarray}
%  \label{eq:fld_radiation}
%  \partial_t E_r + \frac1a \div \left(E_r \myvec{v}_b\right) &=& 
%    \div\left(D\grad E_r\right) -
%    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
%  \label{eq:fld_heating}
%  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \myvec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
%\end{eqnarray}
%where $E_r$ is a grey radiation energy density, $e_c$ is the
%internal energy correction due to photo-heating and chemical cooling,
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  Here, we define the grey radiation energy density
%through first assuming a fixed frequency spectrum, i.e.
%$E_{\nu}(\nu,\myvec{x},t) = \tilde{E}_r(\myvec{x},t) \chi(\nu)$, and then
%defining the integrated quantity
%\begin{equation}
%\label{eq:grey_radiation_energy}
%   E_r(\myvec{x},t) \equiv \int_{\nu_0}^{\infty}
%   E_{\nu}(\nu,\myvec{x},t)\,\mathrm d\nu \  = \ 
%   \tilde{E}_r(\myvec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
%\end{equation}
%In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
%square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
%opacity, $\eta$ is the field of radiation sources, $G$ provides the
%radiation induced photo-heating, $\Lambda$ provides the chemical
%cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_heating}) along with the chemical network 
(Equation~\ref{eq:species_evolution}) is solved
independently of \enzo's hydrodynamics, gravity and dark-matter solvers
(Sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of Equations (\ref{eq:fld_radiation}) and
(\ref{eq:species_evolution}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given timestep to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

### 2013-06-24 numerical-zeus.tex
Changed all "X order" to "x-order" and "finite-difference" and "finite-volume" to "finite difference" and "finite volume", to ensure consistency.

As an alternative to the previous Godunov methods, \enzo\ also includes an implementation of the finite-difference hydrodynamic algorithm employed in the compressible magnetohydrodynamics code \zeus\ \citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} within a multistep (operator split) solution procedure that is fully explicit in time.  This method is formally second order-accurate in space but first order-accurate in time.

@@

The \zeus\ method uses a von Neumann-Richtmyer artificial viscosity to smooth shock discontinuities that may appear in fluid flows and can cause a breakdown of the finite-difference equations.  The artificial viscosity term is added in the source terms as:
\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \vecv - \textbf{Q} : \grad \vecv, 
\end{eqnarray}
Here \textbf{Q} is the artificial viscosity stress tensor, which we take to be diagonal with on-axis terms given by $l^2 \rho (\partial v / \partial x)^2$ as proposed by von Neumann \& Richtmyer.  The length scale $l$ determines the width of shocks and is typically a few times the cell spacing.

@@

Next, we examine the transport step, which is conservative.  Once again, we dimensionally split the equations and present only the one-dimensional version.  The finite-difference equations actually solved are:
\begin{equation}
\rho_j^{n+d} = \rho_j^{n} - \frac{\Delta t}{\Delta x} (v^{n+c}_{j+1/2} \rho^{*}_{j+1/2} - v^{n+c}_{j-1/2} \rho^{*}_{j-1/2} )
\end{equation}
Here $\rho^*_j$ is the correctly upwinded value of $\rho$ evaluated at the cell-face corresponding to $v_j$, making $\rho^*_j v_j$ the mass flux at the cell boundary and guaranteeing mass conservation.   This requires interpolating each cell-centered quantity to the cell edge.  As recommended in \citet{Stone92a}, we use the second-order van Leer scheme, which uses piecewise linear functions.  These are given by equations (48) and (49) of \citet{Stone92a}.  The transport steps for the other variables are similar.  Note that we advect the specific energy and specific momenta using the mass flux, as dictated by the principle of consistent transport.  This requires appropriate averaging for the momenta in the perpendicular directions as outlined in equations (57)-(72) of \citet{Stone92a}.

>>>>

As an alternative to the previous Godunov methods, \enzo\ also includes an implementation of the finite difference hydrodynamic algorithm employed in the compressible magnetohydrodynamics code \zeus\ \citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} within a multistep (operator split) solution procedure that is fully explicit in time.  This method is formally second-order-accurate in space but first-order-accurate in time.

@@

The \zeus\ method uses a von Neumann-Richtmyer artificial viscosity to smooth shock discontinuities that may appear in fluid flows and can cause a breakdown of the finite difference equations.  The artificial viscosity term is added in the source terms as:
\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \vecv - \textbf{Q} : \grad \vecv, 
\end{eqnarray}
Here \textbf{Q} is the artificial viscosity stress tensor, which we take to be diagonal with on-axis terms given by $l^2 \rho (\partial v / \partial x)^2$ as proposed by von Neumann \& Richtmyer.  The length scale $l$ determines the width of shocks and is typically a few times the cell spacing.

@@

Next, we examine the transport step, which is conservative.  Once again, we dimensionally split the equations and present only the one-dimensional version.  The finite difference equations actually solved are:
\begin{equation}
\rho_j^{n+d} = \rho_j^{n} - \frac{\Delta t}{\Delta x} (v^{n+c}_{j+1/2} \rho^{*}_{j+1/2} - v^{n+c}_{j-1/2} \rho^{*}_{j-1/2} )
\end{equation}
Here $\rho^*_j$ is the correctly upwinded value of $\rho$ evaluated at the cell-face corresponding to $v_j$, making $\rho^*_j v_j$ the mass flux at the cell boundary and guaranteeing mass conservation.   This requires interpolating each cell-centered quantity to the cell edge.  As recommended in \citet{Stone92a}, we use the second-order van Leer scheme, which uses piecewise linear functions.  These are given by equations (48) and (49) of \citet{Stone92a}.  The transport steps for the other variables are similar.  Note that we advect the specific energy and specific momenta using the mass flux, as dictated by the principle of consistent transport.  This requires appropriate averaging for the momenta in the perpendicular directions as outlined in equations (57)-(72) of \citet{Stone92a}.
